I don't know if projection is categorical. \\ Did \citealt{mandelkern-etal2020} discover that it is?

### 2023-11-16 writing/naturalness-paper/paper.tex
updated naturalness paper

\citealt[497]{mandelkern-etal2020} urged us to ``think twice before embracing a notion of [\ldots] projection that is gradient based on results from inference tasks alone'', advocating instead for a view of projection that is categorical. In this section, we return to the question of whether projection is a categorical or a gradient property of content. In addressing this question, it is important to distinguish between the projection of {\em content} and the projection of {\em utterance content}. The latter concerns the strength of the projection inference that a particular interpreter draws from a particular utterance of a sentence in which a content is contributed in the scope of an entailment-canceling operator. The empirical evidence pointed to in \S\ref{s1} shows that different interpreters draw projection inferences of varying strength to, for instance, the CC that Julian dances salsa in utterances of {\em Did Cole discover that Julian dances salsa?} In other words, it is uncontroversial that projection is a gradient property {\em utterance content}, not a categorical one. It is gradient because the projection of utterance content is sensitive to several factors on which  interpreters differ, including interpreters' prior beliefs about the speaker or about the utterance content, interpreters' perception and interpretation of the prosodic realization of the utterances, and interpreters' assumptions about the question under discussion addressed by the utterance (e.g., \citealt{djaerv-bacovcin2020,degen-tonhauser-openmind,mahler2020,mahler-thesis,tonhauser-salt26,tbd-variability,tonhauser-etal-sub23}).\footnote{Also: noise, inattentiveness, but replication of variation exclude this.}

What is an open question, however, is whether projection is a gradient or a categorical property of {\em content}, where a content is, for instance, the CC of {\em discover} or the pre-state content of {\em stop}. This question amounts to asking whether the observed projection variability (that is attributable at least to some extent to interpreters) is derived from underlying representations that are categorical or underlying representations that are not categorical. Two aspects of the underlying representations we consider here are speaker belief and lexical entries. A speaker who wants to communicate a content could have a belief in the content that is categorical (that is, they either believe that is is true or they are uncertain) or or that is gradient (in which case they believe the truth to some degree). Likewise, the expressions associated with the content (like {\em discover} or {\em stop}) could have lexical entries that either license a projection inference or not (categorical) or their lexical entries are such that whatever is specified may license a projection inference depending on other factors in the uttered sentence (gradient). This means that there are four hypotheses for any given content:

@@

Some of this hypothesis space has already been advocated for in the literature. Analyses like \citealt{heim83,vds92} and \citealt{potts05} align with hypotheses 1 and 3, as they assume that some expression (triggers of presuppositions or conventional implicatures) are conventionally specified to trigger projection. While these analyses do not specify whether speaker belief is categorical or gradient, one could assume, on the former view (hypothesis 1), that a speaker who believes that a content is true chooses an expression that triggers projection, whereas a speaker who is uncertain about a content chooses an expression that does not trigger projection. On the latter view (hypothesis 3), a speaker believes content to be true to some degree and chooses an expression that either triggers projection or not, depending on what comes closest to their degree of belief. On this view, interpreters might draw projection inferences of different strengths due to uncertainty about the degree of the speaker's belief.

Analyses like those of \citealt{abrusan2011,abrusan2016,abusch02,abusch10,romoli2015} and \citealt{brst-salt10,simons-etal2017} could be taken to align with hypotheses 2 and 4. On some of these analyses, some expressions specify alternatives as part of their lexical entries and this specification of alternatives may, depending on other factors, give lead to a projection inference (e.g., \citealt{abusch02,abusch10,romoli2015}). On other analyses, projection inferences arise from entailments in particular utterances (e.g., \citealt{abrusan2011,abrusan2016,brst-salt10,simons-etal2017}). Crucially, while these analyses differ in many ways, they all share the assumption that the lexical entries of expressions do not categorically trigger (or not trigger) projection. Rather, whether a projection inference is licensed depends on a combination of the lexical entries and other utterance factors.

{\bf MOVE EARLIER: Any of these analyses must be evaluated against the observed projection variability, of which there are two types, namely variability within a given content (e.g., the CC of {\em discover}) as well as variability between different contents (e.g., comparing the CC of {\em discover} and the CC of {\em know}).}

While projection variability is solely due to utterance interpretation on hypothesis 1, it is due to both utterance interpretation and to speaker belief being gradient on hypothesis 3. 
On these analyses, then, the observed projection variability...

\citealt[498f.]{tbd-variability} suggested that projection variability could be due to speaker belief being gradient, as a consequence of which an interpreter's  projection inference is gradient (hypotheses 1, 3, 5, 7). This position still allows for lexical representations being categorical or gradient, and utterance interpretation introducing variability or not. A second position entertained in \citealt[498f.]{tbd-variability} is that speaker belief is categorical and that projection variability ``arises from the [interpreter's] uncertainty about whether the speaker is committed'', that is, as part of the utterance interpretation (hypotheses 2, 6).

p.518: the projectivity of projective content is also influenced by the expression that the projective content is associated with

Recall that \citealt[498]{tbd-variability} proposed that projection is a gradient property of content rather than a categorical one after observing variation in the projection of content in their experiments. Above, we suggested that the results of \citealt{degen-tonhauser-language}, some of which are shown in Fig.~\ref{fig:dt1a}, provide further support for this proposal. \citealt{mandelkern-etal2020}, on the other hand, warned us against ``embracing a notion of presupposition projection that is gradient based on results from inference tasks alone'' (p.497). Contrary to what \citealt{mandelkern-etal2020} led us to expect, the results of our experiment do not provide empirical support for their claim that (presupposition) projection is categorical.

What would constitute empirical evidence that projection is categorical? One type of evidence would be the results of behavioral experiments with a broad range of contents that suggest a nonarbitrary division of contents into ones that typically project and ones that typically don't. Such results does not currently exist, as noted in \citealt{degen-tonhauser-language} and this paper. Another type of evidence could be provided by computational models that evaluate whether the data collected in behavioral experiments is better captured by analyses under which projection is categorical or analyses under which it is gradient.

\item Typically, researchers ask this because they wonder about the linguistic representation we want to assume in our models. Simple picture:

triggers introduce context requirement and hence content typically projects

nontriggers do not introduce this and hence exhibit projection to varying degrees

\item More complex picture, already represented by contemporary analyses:

+SCF: introduce anaphoric requirement/felicity requirement

soft triggers: lexical meaning typically gives rise to projection inference

nontriggers: lexical meaning may give rise to projection inference

\item Contemporary presupposition analyses (e.g., \citealt{heim83,vds92,abrusan2011,karttunen2016,simons-etal2017}) predict neither the heterogeneity of the naturalness ratings of factive predicates nor their sensitivity to the prior probability of the CC (though see \citealt{schlenker2021} for the latter). These results, which mirror those based on inference ratings, suggest that projection analyses must incorporate more fine-grained distinctions between attitude predicates and the systematic effect of prior probabilities on projection.

>>>>

\citealt[497]{mandelkern-etal2020} urged us to ``think twice before embracing a notion of [\ldots] projection that is gradient based on results from inference tasks alone'', advocating instead for a view of projection that is categorical. In this section, we turn to the question of whether projection is a categorical or a gradient property of content. Before addressing the question, we first consider what it is (not) about.

What the question is not about is the projection of {\em utterance content}. What we mean by that is the strength of the projection inference that a particular interpreter draws from a particular utterance of a sentence in which a content is contributed in the scope of an entailment-canceling operator. The empirical evidence pointed to in \S\ref{s1} shows that different interpreters draw projection inferences of varying strength. For instance, as shown in Fig.~\ref{fig:dt1a-JULIAN}, the 266 participants of \citepos{degen-tonhauser-language} Exp.~1a drew projection inferences of varying strength for the CC that Julian dances salsa from utterances of the interrogative {\em Did Scott discover that Julian dances salsa?} It is uncontroversial that this utterance content variability is gradient, not categorical. It is gradient because the projection of utterance content is sensitive to several factors on which interpreters differ, including their prior beliefs about the speaker or about the utterance content, their perception and interpretation of the prosodic realization of the utterances, and their assumptions about the question under discussion addressed by the utterance (e.g., \citealt{djaerv-bacovcin2020,degen-tonhauser-openmind,mahler2020,mahler-thesis,tonhauser-salt26,tbd-variability,tonhauser-etal-sub23}).\footnote{Also: noise, inattentiveness, but replication of variation exclude this.}

\begin{figure}[h!]
\centering
\includegraphics[width=.8\textwidth]{../../results/main/13explicitIgnorance/graphs/mean-certainty-by-predicateType-JULIAN}
\caption{Participants' certainty ratings (measuring projection) of the CC that Julian dances salsa for the 20 \color{orange}factive \color{black} and \color{green}nonfactive \color{black} predicates investigated in Exp.~1a of \citealt{degen-tonhauser-language} (adapted by collapsing the nonfactive predicates into a single category).}\label{fig:dt1a-JULIAN}
\end{figure}

What the question about whether projection is gradient or categorical is about is {\em content}, where a content is, for instance, the CC of {\em discover}, the CC of {\em inform}, or the pre-state content of {\em stop}. Asking whether projection is gradient or categorical at this level of abstraction (that is, abstracting away from variation attributable to interpreter variability) amounts to asking whether the observed by-content variability (for instance that the CC of {\em discover} is less projective than the CC of {\em inform}; see Fig.~\ref{fig:dt1a}) is derived from underlying representations that are categorical or underlying representations that are not categorical. Two aspects of the underlying representations we consider here are speaker belief and lexical entries. A speaker who wants to communicate a content might have a belief in the content that is categorical (that is, they either believe that is is true or they are uncertain) or one that is gradient (in which case they believe that the content is true to some degree).\footnote{\citealt[498f.]{tbd-variability} already considered that speaker belief could be gradient or categorical.} Likewise, the expressions associated with the content (like {\em discover} or {\em stop}) could have lexical entries that either license a projection inference or not (categorical) or their lexical entries could be such that whatever is specified may license a projection inference depending on other factors in the uttered sentence (gradient). This means that there are four hypotheses for any given content:

@@

Some of this hypothesis space has already been advocated for in the literature. Analyses like \citealt{heim83,vds92} and \citealt{potts05} align with hypotheses 1 and 3, as they assume that some expression (triggers of presuppositions or conventional implicatures) are conventionally specified to trigger projection. While these analyses do not specify whether speaker belief is categorical or gradient, one could assume, on the former view (hypothesis 1), that a speaker who believes that a content is true chooses an expression that triggers projection, whereas a speaker who is uncertain about a content chooses an expression that does not trigger projection. On the latter view (hypothesis 3), a speaker believes content to be true to some degree and chooses an expression that either triggers projection or not, depending on what comes closest to their degree of belief. On either view, interpreters might draw projection inferences of different strengths due to uncertainty about (the degree of) the speaker's belief.

Analyses like those of \citealt{abrusan2011,abrusan2016,abusch02,abusch10,romoli2015} and \citealt{brst-salt10,simons-etal2017} could be taken to align with hypotheses 2 and 4. On some of these analyses, some expressions specify alternatives as part of their lexical entries and this specification of alternatives can, depending on other factors, give lead to a projection inference (e.g., \citealt{abusch02,abusch10,romoli2015}). On other analyses, entailed content may project in particular utterances  (e.g., \citealt{abrusan2011,abrusan2016,brst-salt10,simons-etal2017}). Crucially, while these analyses differ in many ways, they all share the assumption that the lexical entries of expressions do not categorically trigger (or not trigger) projection. Rather, whether a projection inference is licensed depends on a combination of the lexical entries and other utterance factors. These analyses, too, don't specify whether speaker belief is categorical or gradient.

To evaluate which of the four hypotheses in Table \ref{t:hyp} is appropriate for any given content, we have to consider the observed by-content variability. An extreme position is to assume for every projective content that speaker belief is categorical and triggered by conventional specification. To capture the observed by-content variability, one would then have to assume that it arises during utterance interpretation and affects different expression/content pairs differently. That does not seem plausible. A more plausible position seems to us to be one on which each of the four hypotheses is correct but for different sets of expression/content pairs. For instance, hypothesis 1 might be appropriate  the existence requirement of pronouns and {\em too} (expression/content pairs associated with a strong contextual felicity requirement), hypothesis 2 for the pre-state of {\em stop}, and hypothesis 4 for the CC of {\em inform}. The observed by-content variability would then be attributable to a combination of factors, with expression/content pairs differing on the factors and their relative weights.

There is, to date, no projection analysis that can predict the observed by-content variability (for discussions, see \citealt{tbd-variability,degen-tonhauser-language}). What seems clear, given the observed by-content variability, is that the question of whether projection is a gradient or a categorical property of content does not have a single answer for all projective content. What the source of the variability is must be decided in future research that carefully compares...
